---
title: "Tidying and Merging Data"
author: "Glenn Williams"
date-format: "YYYY-MM-DD"
date: "`r format(Sys.time(), '%Y %B, %d')`"
execute:
  echo: true
format: 
  revealjs:
    theme: [styles.scss]
    preview-links: auto
    transition: slide
    min-scale: 0.1
editor: visual
---

## Overview

::: r-fit-text
Today's session is on tidying and merging data:

-   Reading in data multiple data sets at once using the `read_*()` family of functions, e.g. `read_csv()`.

-   Binding columns and rows to your data set using `bind_cols()` and `bind_rows()`.

-   Understanding wide and long data formats and how to reshape your data to either format using `pivot_longer()` and `pivot_wider()`.

-   Joining data together from multiple tables using mutating joins such as `inner_join()`, `left_join()`, `right_join()`, and `full_join()`.

-   Filtering to rows in one column based on presence or absence of matches in another using `semi_join()` and `anti_join()`.

-   Perform set operations and remove duplicates across compatible data sources (i.e. with the same columns) using `intersect()`, `union()`, `setdiff()`, and `distinct()`.
:::

# Getting Started

-   Go to <https://github.com/gpwilliams/ds-psych_course>.

-   Click `Code` \> Download ZIP.

-   Unzip the files.

-   Open the file `ds-psych_course.RProj`

-   Create a Quarto document and save it within `04_tidying-and-merging-data`. Name it anything you like.

You can copy the code form here on out and everything should work.

## Loading Packages

Load our essential packages, `tidyverse` for data wrangling and presentation and `here` for working with file paths.

```{r}
library(tidyverse)
library(here)
```

# Reading Data

-   Often, you might work with programs such as PsychoPy which output individual files for each participant from an experiment.

-   With some functions you might need to use a loop or the `map_*()` family of functions to iterate over a process of reading them in one by one before binding them to one `tibble`.

-   As of `readr` 2.0.0 you can now provide just a list of files to the `read_*()` family of functions.

## Reading Data

::: r-fit-text
We must define a file path to the folder in which all the files are stored.

```{r}
files <- list.files(
  path = here("data", "tidying-and-merging", "stroop"),
  recursive = TRUE, # check subdirectories?
  pattern = "\\.csv$", # file ending can only be .csv
  full.names = TRUE # we need the full path
)
```

Print `files` object to see this list in full.

```{r}
#| message: false
stroop <- read_csv(files)
stroop
```
:::

# Pivoting

::: r-fit-text
-   Many of the `tidyverse` functions and statistical packages expect data to be in a long/**tidy** format:

    -   Each column represents a variable.
    -   Each row represents an observation.

-   More rarely, functions expect data in a wide format:

    -   Each row represents a case (e.g. participant).
    -   Levels of variables are spread across columns.

*Moving between the two is often necessary, even if you don't intend to stay within the R ecosystem.*
:::

# Pivoting

## Pivoting to a wider format

We use `pivot_wider()` to take data from a long format and make it wide.

It takes as arguments:

-   `data`: data source (e.g. your tibble) from which to pivot.
-   `names_from`: column containing the IDs that will become column headings.
-   `values_from`: column containing the values that will be split between columns.

*We can specify numerous `names_from` or `values_from` with `c()`.*

## `pivot_wider()`

```{r}
stroop_wide <- stroop |> 
  pivot_wider(
    names_from = stroop,
    values_from = rt
  )
stroop_wide
```

## Pivoting to a longer format

We use `pivot_longer()` to take data from a wide format and make it long.

It takes as arguments:

-   `data`: data source (e.g. your tibble) from which to pivot.
-   `cols`: the columns to pivot into a longer format.
-   `names_to`: the new column name for identifying e.g. levels of a factor.
-   `values_to`: the new column name for identifying e.g. the dependent variable.

## `pivot_longer()`

```{r}
stroop_long <- stroop_wide |> 
  pivot_longer(
    cols = c(Incongruent, Neutral),
    names_to = "condition",
    values_to = "rt"
  )
stroop_long
```

# Joining

Often we want to combine multiple data sources into one table. We can do this to:

-   Bind: append rows or columns to a `tibble`.
-   Mutate: change a single table by combining columns and rows from another table based on shared identifiers.
-   Filter: Subset a single table by using a reference to another table based on shared identifiers.

# Binding Joins

## Loading more data

We'll load some new data sets for this example.

-   `demo`: some demographics for each participant in the original Stroop data.
-   `reading-coder-1`: some details of Language expertise for each participant in the original Stroop data.
-   `stroop_new`: a new data set of participants taking part in the Stroop task.

```{r}
#| message: false
demo <- read_csv(here("data", "tidying-and-merging", "stroop_demographics.csv"))
reading_coder_1 <- read_csv(here("data", "tidying-and-merging", "reading_coder-1.csv"))
stroop_new <- read_csv(here("data", "tidying-and-merging", "stroop_new.csv"))
```

## Binding Columns

Check out the `demo` and `reading_coder_1` data sets.

::: columns
::: {.column width="40%"}
```{r}
demo
```
:::

::: {.column width="60%"}
```{r}
reading_coder_1
```
:::
:::

We can use `bind_cols()` to append the columns of the new data set to the original one.

## `bind_cols()`

::: r-fit-text

::: callout-note
This only works if the number of rows matches across data sets.
:::

```{r}
bind_cols(demo, reading_coder_1)
```

::: callout-warning
**Why do we get multiple ID cols?** Shared column headings are repeated. This is safe behaviour so we can check they align. We really should avoid binding columns unless we are certain the rows are perfectly aligned.
:::

:::

## Binding rows

Check out the `stroop` and `stroop_new` data sets.

::: columns
::: {.column width="50%"}
```{r}
stroop
```
:::

::: {.column width="50%"}
```{r}
stroop_new
```
:::
:::

## `bind_rows()`

We can use `bind_rows()` to append the rows from the new data set to the original one.

```{r}
bind_rows(stroop_new, stroop)
```

## Identifying data in `bind_rows()`

::: r-fit-text
-   `bind_rows()` has the optional argument, `.id` to give a column heading identifying which data set observations come from.

-   Give a string vector to define the column heading.

```{r}
bind_rows(stroop_new, stroop, .id = "data_set")
```
:::

# Mutating Joins

If we want to have a single data set that combines values from another data set using a shared identifier to combine them, we can use a **mutating** join.

-   **Outer Joins**: Keep observations that with identifiers that appear in at least one of the data sets.
-   **Inner Joins**: Only keep observations in one data set that has matching identifiers in another data set.

## Mutating Joins

All take the arguments:

-   `x`: the left data set (e.g. a tibble)
-   `y`: the right data set (e.g. a tibble)
-   `by`: the column heading to act as a key/identifier across data sets passed as a string vector. This is **optional but encouraged** to not rely on defaults.

## Full Joins (an Outer Join)

-   Keep all observations in `x` and `y`.
-   Notice that we get `NA`s for values that weren't present in one of the data sets.

```{r}
full_join(x = stroop, y = demo, by = "subject_id")
```

## Left Join (an Outer Join)

-   Keep all observations in `x` and match identifiers for these that occur in `y`.
-   `NA`s are only introduced to values that originally existed in `x` but don't have a value in `y`.

```{r}
left_join(x = stroop, y = demo, by = "subject_id")
```

## Right Join (an Outer Join)

-   Keep all observations in `y` and match identifiers for these that occur in `x`.
-   `NA`s are only introduced to values that originally existed in `y` but don't have a value in `x`.

```{r}
right_join(x = stroop, y = demo, by = "subject_id")
```

## Inner Join (...an Inner Join 🙄)

-   Only keep observations from `x` that have a matching identifier in `y`.
-   We only get complete cases.

```{r}
inner_join(x = stroop, y = demo, by = "subject_id")
```

When you have multiple matches, there is advanced help here by using the `join_by()` functions. We won't go into this.

# Inconsistent Naming

Inconsistent naming makes working with joins more difficult, but not impossible. Either:

-   **Rename the headings** so they're consistent across data sets. (We already know how to use `rename()`)
-   **Define how identifiers** map onto one another in the join call.

## Inconsistent Naming

First, let's break the naming conventions.

```{r}
demo_2 <- rename(demo, participant = subject_id)
demo_2
```

This won't work as the two data sets don't have a shared identifier:

```{r}
#| eval: false
full_join(stroop, demo_2)
```

## Fixing Inconsistent Naming

All joins can pass the `join_by()` function to `by` to allow us to specify how column identifiers are mapped onto one another.

```{r}
full_join(
  stroop, 
  demo_2, 
  by = join_by(subject_id == participant)
)
```

## Fixing Inconsistent Naming

By default it keeps the first name and merges the values, this can be fixed by using `keep = TRUE`.

```{r}
full_join(
  stroop, 
  demo_2, 
  by = join_by(subject_id == participant),
  keep = TRUE
)
```

# Filtering Joins

-   Often we might want to filter rows out of one data set based on values in another one, e.g. if we have a list of people who need to be excluded or dropped out of our study.

-   Filtering Joins filter rows from `x` based on the presence or absence of matches to identifiers in `y`.

## Semi-Join

-   A semi join **keeps** all rows with an identifier in `x` that have a match in `y`.
-   This works like `inner_join()` but doesn't append the columns from `y`.

```{r}
semi_join(stroop, demo, by = "subject_id")
```

## Anti-Join

-   A semi join **removes** all rows with an identifier in `x` that have a match in `y`.
-   This works like the **inverse** of `inner_join()` but doesn't append the columns from `y`.

```{r}
anti_join(x = stroop, y = demo, by = "subject_id")
```

## Anti-Join

Which data set you define as `x` and `y` has important ramifications. Compare this to the previous result.

```{r}
anti_join(demo, stroop, by = "subject_id")
```

# Set Operations

## Load more data

We'll load up some more data to help us see some additional cases where joins are useful.

-   `reading_coder_2`: some details of Language expertise for each participant in the original Stroop data, this time rated by another coder such that values may be inconsistent with `reading_coder_1`.
-   `reading_dupe`: the same data as `reading_coder_1` but with a duplicate row.

```{r}
#| message: false
reading_coder_2 <- read_csv(here("data", "tidying-and-merging", "reading_coder-2.csv"))
reading_dupe <- read_csv(here("data", "tidying-and-merging", "reading_dupe.csv"))
```

## Union

Check out the `reading_coder_1` and `stroop_new` data sets.

::: columns
::: {.column width="50%"}
```{r}
reading_coder_1
```
:::

::: {.column width="50%"}
```{r}
reading_coder_2
```
:::
:::

## `union()`

Find all rows in either `x` or `y`, excluding duplicates.

```{r}
union(reading_coder_1, reading_coder_2)
```

## Intersect

Find all rows that occur in **both** `x` and `y`.

```{r}
intersect(reading_coder_1, reading_coder_2) 
```

## Set Diff

Find all rows in `x` that **aren't** in `y`.

```{r}
setdiff(reading_coder_1, reading_coder_2) 
```

## Set Diff

Check out the `reading_coder_1` and `reading_dupe` data sets.

::: columns
::: {.column width="50%"}
```{r}
reading_coder_1
```
:::

::: {.column width="50%"}
```{r}
reading_dupe
```
:::
:::

## `union()`

```{r}
union(reading_coder_1, reading_dupe)
```

## Working with duplicates

What if we want to exclude duplicate rows from a single data set without using a key from another data set? Just use `distinct()`!

```{r}
distinct(reading_dupe)
```

## Exercises

Please complete the exercises at <https://github.com/gpwilliams/ds-psych_course>.
